---
title: "Design for Emergence"
output: html_notebook
---

## Introduction

MBITES doesn't link eggs laid by mosquitoes to the number of imagos. It could, but it doesn't yet. The current version draws from a Poisson-distributed rate of emergence. This document lays out a way to calculate how mosquito egg-laying produces imagos.

We start with a description of the current code in order to see the order of events and where, in the code, each step is calculated. Then we can make a plan to teach the code to do what we want.

We have a simple goal. The eggs get laid and have a daily survival.

$$
L_{t+1} = \mbox{eggs}_t + \mbox{survive}_t L_t - \mbox{mature}_t L_t
$$

Imagos are Poisson-distributed from those egg counts.
$$
\mbox{Imago}_t = \mbox{Poisson}(\mbox{mature}_t L_t)
$$

## Sequence of events

The context for emergence is that it is calculated during the mosquito's portion of the time step. The high-level function call is the bout.
```{r, eval=FALSE}
mbites_oneBout <- function(){

  # launch at the previously scheduled launch time
  private$tNow = private$tNext

  # launch and try
  if(private$search){
    self$attempt_search() # search
    private$searchNow = TRUE
  } else {
    # attempt an action
    switch(private$state,
      B = {self$attempt_B()},
      O = {self$attempt_O()},
      M = {self$attempt_M()},
      S = {self$attempt_S()},
      {stop("mosquito ",private$id," calling illegal behavioral state: ",private$state,"\n")}
    )
    private$searchNow = FALSE
  }

  # land
  self$restingSpot() # MBITES-Resting.R

  # rest
  self$updateState() # MBITES-Bout.R

  # log history
  if(private$alive){
    self$trackHistory()
  }
}
```

Within the `updateState()` call, it computes oogenesis and other steps.
```{r, eval=FALSE}
mbites_updateState <- function(){
  # only do this if you are alive
  if(private$alive){

    self$energetics()    # MBITES-Energetics.R
    self$survival()      # MBITES-Survival.R

    # check again for being alive because can be killed in survival; don't want zombie mosquitoes preforming actions
    if(private$alive){

      if(private$bloodfed){
        self$Oogenesis() # MBITES-Oogenesis.R
      }
      # This can change the bout state from B to O.
      self$checkEggMaturation() # MBITES-Oogenesis.R

      # The states in priority order
      if(private$starved){
        private$state = "S"
      } else {
        if(private$gravid){
          # Refeeding conditionally moves the bout state to B.
          self$checkRefeed()  # MBITES-Oogenesis.R
        } else {
          private$state = "B"
        }
      }

      # if there are no resources of the required type present, set
      # search = TRUE
      self$checkForResources()

      # sample time to next launch based on current state (before it is changed from energetics, survival, egg maturation, or reeding)
      self$timing() # MBITES-Timing.R

      # check time-dependent events
      self$findSwarm()
      self$checkEstivation()

      # call pathogen dynamics last so that EIP uses the tNext (time of next launch)
      # to increment its incubation period now. this ensures that at the start of a bout,
      # the pathogen is referencing the correct day.
      self$pathogenDynamics() # PATHOGEN-XX.R
    }
  }
}
```

If the mosquito isn't in the B bout, for bloodfeed, it can switch to the O bout, for Oviposition.
```{R, eval=FALSE}
mbites_attempt_O <- function(){

  # bout success
  if(runif(1) < MBITES:::Parameters$get_O_succeed()){

    # sample resources
    self$chooseHabitat() # MBITES-Oviposition.R

    # check habitat type
    if(private$habitatID > 0L){
      # This should put eggs into the EggQ where emergence will find them.
      self$layEggs() # MBITES-Oviposition.R
    } else if(private$habitatID == -1L){
      # self$ovitrap()
    } else {
      stop("illegal habitatID value")
    }
  }

  # bout failure
  if(private$batch > 0){
    # if i did not succeed my bout, increment failure counter
    private$boutFail = private$boutFail + 1L
  } else {
    # reset flags
    private$boutFail = 0L
  }
}
```

We can see these as the following steps, in time, for the mosquito.

1. Bloodfeed - Part of the feeding bout in the time step.

2. Oogenesis

   a. Refeeding is part of oogenesis algorithm. Can be egg batch refeeding or bloodmeal refeeding.

   b. Egg batch size is decided here. Can be Gaussian-distributed or bloodmeal-dependent.

   c. Two types of oogenesis: Either enter gravid state after egg maturation time has passed, or enter gravid state after egg provision is filled.

3. Oviposition - picks the site. Puts eggs into the EggQ at the aquatic resource for that site. This code is in `LANDSCAPE-Emerge.R`.

4. Emergence - Draws from emergence.

5. Imagos


## Structure

* The mosquito has

  * `gravid` whether it is gravid to oviposit.
  * `energyPreG` the pre-gonotrophic energy requirement.
  * `bloodfed` boolean.
  * `batch` whether it has fed on blood this bout. Set during oogenesis.
  * `eggT` size of the egg batch. Set during oogenesis.
  * `eggP` time the egg batch will be ready. Set during oogenesis.
  * `bmSize` size of the bloodmeal.

* The aqua resource has

  * `eggQ` the eggs laid.
  * `imagoQ` imagos.
