---
title: "Design for Emergence"
output: html_notebook
---

## Introduction

MBITES doesn't link eggs laid by mosquitoes to the number of imagos. It could, but it doesn't yet. The current version draws from a Poisson-distributed rate of emergence. This document lays out a way to calculate how mosquito egg-laying produces imagos.

We start with a description of the current code in order to see the order of events and where, in the code, each step is calculated. Then we can make a plan to teach the code to do what we want.

We have a simple goal. The eggs become larvae and have daily survival.

$$
L_{t+1} = \mbox{eggs}_t + \mbox{survive}_t L_t - \mbox{mature}_t L_t
$$

Imagos are Poisson-distributed from those larva counts.
$$
\mbox{Imago}_t = \mbox{Poisson}(\mbox{mature}_t L_t)
$$
Those are our two goals.

## Sequence of events in the code

The MBITES simulation has a daily time step, within which each mosquito performs bouts. The following code executes the next bout, and emergence is a background action of the bout, so it happens during `updateState()` near the bottom of the function.
```{r, eval=FALSE}
mbites_oneBout <- function(){

  # launch at the previously scheduled launch time
  private$tNow = private$tNext

  # launch and try
  if(private$search){
    self$attempt_search() # search
    private$searchNow = TRUE
  } else {
    # attempt an action
    switch(private$state,
      B = {self$attempt_B()},
      O = {self$attempt_O()},
      M = {self$attempt_M()},
      S = {self$attempt_S()},
      {stop("mosquito ",private$id," calling illegal behavioral state: ",private$state,"\n")}
    )
    private$searchNow = FALSE
  }

  # land
  self$restingSpot() # MBITES-Resting.R

  # rest
  self$updateState() # MBITES-Bout.R

  # log history
  if(private$alive){
    self$trackHistory()
  }
}
```

Within the `updateState()` call, it computes oogenesis and other steps.
```{r, eval=FALSE}
mbites_updateState <- function(){
  # only do this if you are alive
  if(private$alive){

    self$energetics()    # MBITES-Energetics.R
    self$survival()      # MBITES-Survival.R

    # check again for being alive because can be killed in survival; don't want zombie mosquitoes preforming actions
    if(private$alive){

      if(private$bloodfed){
        self$Oogenesis() # MBITES-Oogenesis.R
      }
      # This can change the bout state from B to O.
      self$checkEggMaturation() # MBITES-Oogenesis.R

      # The states in priority order
      if(private$starved){
        private$state = "S"
      } else {
        if(private$gravid){
          # Refeeding conditionally moves the bout state to B.
          self$checkRefeed()  # MBITES-Oogenesis.R
        } else {
          private$state = "B"
        }
      }

      # if there are no resources of the required type present, set
      # search = TRUE
      self$checkForResources()

      # sample time to next launch based on current state (before it is changed from energetics, survival, egg maturation, or reeding)
      self$timing() # MBITES-Timing.R

      # check time-dependent events
      self$findSwarm()
      self$checkEstivation()

      # call pathogen dynamics last so that EIP uses the tNext (time of next launch)
      # to increment its incubation period now. this ensures that at the start of a bout,
      # the pathogen is referencing the correct day.
      self$pathogenDynamics() # PATHOGEN-XX.R
    }
  }
}
```

If the mosquito isn't in the B bout, for bloodfeed, it can switch to the O bout, for Oviposition.
```{R, eval=FALSE}
mbites_attempt_O <- function(){

  # bout success
  if(runif(1) < MBITES:::Parameters$get_O_succeed()){

    # sample resources
    self$chooseHabitat() # MBITES-Oviposition.R

    # check habitat type
    if(private$habitatID > 0L){
      # This should put eggs into the EggQ where emergence will find them.
      self$layEggs() # MBITES-Oviposition.R
    } else if(private$habitatID == -1L){
      # self$ovitrap()
    } else {
      stop("illegal habitatID value")
    }
  }

  # bout failure
  if(private$batch > 0){
    # if i did not succeed my bout, increment failure counter
    private$boutFail = private$boutFail + 1L
  } else {
    # reset flags
    private$boutFail = 0L
  }
}
```

There is a second place in the code that that makes the daily timestep happen. The code is, at its top level, a set of `HashMap` objects. The main responsibility of a `HashMap` object is to enable a quick call to `apply`, that is faster than `lapply` because it doesn't accept return values. In this case, it calls `oneDay_AquaticEcology_Site`, which is in `LANDSCAPE-Site.R:141`. This calls `one_day()` in emergence and `push_imago()` in oviposition.


## Sequence of events in time

We can see these as the following steps, in time, for the mosquito.

1. Bloodfeed - Part of the feeding bout in the time step.

2. Oogenesis

   a. Refeeding is part of oogenesis algorithm. Can be egg batch refeeding or bloodmeal refeeding.

   b. Egg batch size is decided here. Can be Gaussian-distributed or bloodmeal-dependent.

   c. Two types of oogenesis: Either enter gravid state after egg maturation time has passed, or enter gravid state after egg provision is filled.

3. Oviposition - picks the site. Puts eggs into the EggQ at the aquatic resource for that site. This code is in `LANDSCAPE-Emerge.R`.

4. Emergence - Draws from emergence.

5. Imagos


## Structure

* The mosquito has these member variables.

  * `gravid` whether it is gravid to oviposit.
  * `energyPreG` the pre-gonotrophic energy requirement.
  * `bloodfed` boolean.
  * `batch` whether it has fed on blood this bout. Set during oogenesis.
  * `eggT` size of the egg batch. Set during oogenesis.
  * `eggP` time the egg batch will be ready. Set during oogenesis.
  * `bmSize` size of the bloodmeal.

* The aqua-emerge resource has two member variables which represent queues of eggs and imagos.

  * `eggQ` the eggs laid.
  * `imagoQ` imagos.


## How to modify for emergence

### How the code is set up to support variation

1. The oogenesis code, in `MBITES-Oogenesis.R`, works by attaching methods to the mosquito class. It supports variation on initialization of the mosquito, where it chooses which methods to attach. This happens in `MBITES_Setup_Oogenesis()` in the `MBITES-Setup.R` file. The calls to initialize this happen in the user-level script, `sei-simulation.R`, in this case. By the time oogenesis code is done, the mosquito has a batch of eggs in its `eggT` and `eggP` members.

2. The eggs move to the site during oviposition, which is in the `MBITES-Oviposition.R` code. These methods are attached to the mosquito class in the `MBITES_Setup_MBDETES()` function, which is in `MBDETES-Setup.R`. The current code attaches a method which _does not oviposit eggs._ There is, in the same file, an example of a function which does oviposit a batch of eggs. It's called `mbites_layEggs_EL4P()`, and its code inserts eggs into the `eggQ` at an aqua emergence site.

3. The code for emergence is in the `Aqua_Emerge_Resource`, which is a subclass of `Aqua_Resource`. The current version of this class doesn't calculate daily survival of eggs. It also doesn't draw from a Poisson distribution to create imagos, so we would make a different version of `Aqua_Emerge_Resource` in order to match the equations we want.

4. The imagos, too, are part of the `Aqua_Emerge_Resource`, so we would add those calculations in the same class.


### What we can do

We need to insert a larval stage between egg maturation and production of imagos. That is squarely in the Aqua-Emerge resource class. So we have to modify that class or make a different version of it, which makes more sense. There is another problem we could address.

It's difficult to configure the simulation by attaching methods to the mosquito. You have to know what methods it needs and which methods work correctly with each other. For instance, the oogenesis code describes problems mixing-and-matching the refeeding and egg-laying methods. We can see in a user-level script, `sei-simulation.R`, that the user has to calls lots of setup before starting the simulation, which we expect, but some of that setup is about attaching methods to classes, which is unusual and requires more knowledge. So can we make small changes to the current code to make it easier a) to implement this emergence calculation and b) to set up a simulation?

We can solve this problem by continuing to use the same tools, at the egg-level, that we used at the top level of the code. We can create a class structure around eggs and the aquatic stages.

The first two stages are associated with the mosquito and oogenesis. The central object is an egg batch, and its presence interacts with refeeding and terminates in oviposition. We could make the egg batch part of a class that represents oogenesis. Choosing a habitat is a short function that looks at sites. Maybe this can be an if-then statement that switches based on a configuration parameter.

The next two stages are associated with the site. The same `Aqua_Emerge` resource covers both steps, growth of the eggs and holding imagos. Maybe we can keep this the same way and add another version of `Aqua_Emerge` that has the mathematical behaviors we want.


### Development steps

The eggs aren't being created, so the first step is to fix that function in the oviposition step. That's small once you know where to find it.

Create a substitute for the `Aqua_Emerge` class. That's it. We don't have to fix the whole thing right now. If we want to do more, keep in mind that it's much easier to separate steps where you move code from steps where you change features. So it would start with taking existing egg functionality and putting it into its own class.

The eggs and imagos are in a queue data structure. Can we put the larvae in a similar data structure?
